Modal Verification Patterns for Systems

### 2025-06-23 paper.tex
+

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
\todo[inline,color=pink]{Based on the note I took above, I changed the text}
A typical general purpose computers make the memory resources in RAM as \emph{virtually} available. In other words, a program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized} such that it may be \emph{mapped to} a physical RAM location or not. As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS kernels relies \emph{translating}(\emph{mapping}) virtual addresses into physical ones. The conceptual address-translation map is implemented with processor's(CPU) page-table trees as showin in Figure \ref{fig:pagetables2.png}.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}\todo{inline,color=pink}{I am well-aware what counts as as (including self-plagiarism). These were annotated with todos including cautions for changing the text.}
On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables called \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

The entries of each table are 64 bits wide, but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set), and a range of additional bits which can be used to control caching, write-through, and more. This paper will only consider the present bit (0), the read-write bit (1), and the accessed bit (5), which is set to 1 by the hardware every time an entry is used, allowing the OS to determine which pieces of memory are used or unused.

The page tables mapping virtual addresses to physical addresses are managed by the OS. Typically each process has its on virtul-to-physical mappings represented by its own page table, which the OS registers with the CPU by storing it in a specific register (\texttt{cr3}) as part of switching to a new process. Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) is how the OS ensures memory isolation between processes.

If an instruction is executed that accesses a virtual address that either has no mapping, or does not have a mapping permitting the kind of access that was performed (e.g., the instruction was a memory write, but the relevant address range was marked read-only in the relevant page table entry), the hardware triggers a \emph{page fault}, transferring control to a \emph{page fault handler} registered with the hardware by the OS, allowing it to take corrective action (described in the next section).

\paragraph{Virtual Memory Managers with Multiple Address-Spaces}
\label{sec:backgroundonvmm}
\todo{change the text}
Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

>>>>

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
\todo[inline,color=pink]{Based on the note I took above, I changed the text}
A typical general purpose computers make the memory resources in RAM as \emph{virtually} available. In other words, a program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized} such that it may be \emph{mapped to} a physical RAM location or not. As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS kernels relies \emph{translating}(\emph{mapping}) virtual addresses into physical ones. The conceptual address-translation map is implemented with processor's(CPU) page-table trees as shown in Figure \ref{fig:pagetables2.png}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the top indexing into the different levels of the page-table tree. Fabrication of virtual addresses out of limited physical addresses is provided with entries in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. An address translation requires traversal of a series of tables starting from level 4, and it is conventionally referred as a \emph{page-table walk}. The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}\todo{inline,color=pink}{I am well-aware what counts as as (including self-plagiarism). These were annotated with todos including cautions for changing the text.}
\paragraph{Virtual Memory Managers with Multiple Address-Spaces}
\label{sec:backgroundonvmm}
\todo{change the text}
The page tables mapping virtual addresses to physical addresses are managed by the OS. Typically each process has its on virtul-to-physical mappings represented by its own page table, which the OS registers with the CPU by storing it in a specific register (\texttt{cr3}) as part of switching to a new process. Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) is how the OS ensures memory isolation between processes.

Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.
